---
title: "Electiva Activity"
author: "Dana Gabriela Salazar Toro"
date: "`2023-08-29`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This report presents an analysis of flight data using the *n**ycflights13*** package and ***tidyverse*** in R. Various aspects of the flight data, such as delays, times, and punctuality, are explored.

# Load ne cessary libraries

```{r message=FALSE}
library(nycflights13) 
library(tidyverse)
```

In this section, we are loading the libraries needed to perform data analysis. nycflights13 contains the flight dataset and tidyverse is a set of packages that facilitate data manipulation and visualization.

# Load the flights dataset

Here, we are loading the flight data set into the **flights_data** object, using the flights function of the **nycflights13** package.

```{r}
f_d <- nycflights13::flights
```

# 5.2.4 Exercises: Items 1 and 2

This code block solves item 1 of the exercise. We are using the **filter()** function of the tidyverse package to select only the **flights_data** rows where the value in the **arr_delay** (delay on arrival) column is greater than or equal to 2. The result is stored in the **filtered_delayed_flights** object.

```{r}
myDf1 <- filter(f_d, arr_delay >= 2)
```

In this code block:

We made use of the **Knitr** function in order to plot the table taking into account the tail number of the aircraft and the landing delay time:

```{r}
library(knitr)
kable(f_d[1:10,c(12,9)],caption = "ARRIVE DELAY", align = "c")
```

This code block solves item 2 of the exercise. Similarly, we are using the **filter()** function to select the rows where the value in the dest column is equal to "HOU". The result is stored in the **filtered_hou_flights** object.

```{r}
myDf2 <- filter(f_d,dest =="HOU")
```

```{r}
library(knitr)
kable(myDf2[1:10,c(13,14)],caption = "HOUSTON DESTINY", align = "c")
```

# 5.3.1 Exercises: All Items

```{r}
sorted_flights_missing_first <- flights %>%
  arrange(desc(is.na(dep_time)))
```

In this code block, we are performing the following actions:

1.  We take the data set flights and pass it through the **%\>%** operator, which allows us to chain operations.

<!-- -->

2.  We use the **arrange()** function to sort the rows in descending order based on the i**s.na(dep_time)** column. This column is logical and returns TRUE if **dep_time** (departure time) is absent and FALSE if it is not. This means that we are sorting the flights so that those with missing values in **dep_time** appear first.

```{r}
library(knitr)
kable(sorted_flights_missing_first[1:10,c(7,12)],caption = "MISSING DATA FIRST", align = "c")
```

```{r}
most_delayed_flights <- flights %>%
  arrange(desc(arr_delay))
```

In this code block:

1.  We take the flights **dataset** and pass it through the **%\>%** operator.
2.  We use the **arrange()** function to sort the rows in descending order based on the **arr_delay** column. This means that flights with longer arrival delays will appear first.

```{r}
library(knitr)
kable(most_delayed_flights[1:10,c(6,9,12)],caption = "MOST DELAYED FLIGHTS", align = "c")
```

```{r}
fastest_flights_desc <- flights %>%
  mutate(speed = distance / air_time) %>%
  arrange(desc(speed))
```

In this code:

1.  We take the **`flights`** dataset and pass it through the **`%>%`** operator.

2.  We use the **`mutate()`** function to create a new column called **`speed`**. We calculate the speed by dividing the **`distance`** column by the **`air_time`** column. This will give us the speed of each flight.

3.  After creating the **`speed`** column, we use the **`arrange()`** function to sort the rows in descending order based on the **`speed`** column. This means that flights with the highest speed will appear first.

```{r}
library(knitr)
kable(fastest_flights_desc[1:10,c(12,20)],caption = "FASTEST FLIGHTS", align = "c")
```

```{r}
farthest_flights <- flights %>%
  arrange(desc(distance))
```

In this code:

1.  We take the **`flights`** dataset and pass it through the **`%>%`** operator.

2.  We use the **`arrange()`** function to sort the rows in descending order based on the **`distance`** column. This means that flights with the longest distances will appear first.

```{r}
library(knitr)
kable(farthest_flights[1:10,c(12,15)],caption = "FARTHEST FLIGHTS", align = "c")
```

```{r}
closest_flights <- flights %>%
  arrange(distance)
```

In this code:

1.  We take the **`flights`** dataset and pass it through the **`%>%`** operator.

2.  We use the **`arrange()`** function to sort the rows in ascending order based on the **`distance`** column. This means that flights with the shortest distances will appear first.

```{r}
library(knitr)
kable(closest_flights[1:10,c(12,15)],caption = "CLOSEST FLIGHTS", align = "c")
```

# 5.4.1 Exercises: Items 2, 3, and 4

R's **`select()`** function, if you include the name of a variable multiple times, it will appear multiple times in the resulting output. For example:

```{r message=FALSE}
select(flights, dep_time, dep_time)
```

This would include the **`dep_time`** column twice in the resulting output.

The **`any_of()`** function is used to select columns from a dataframe based on a character vector of column names. It's helpful when you have a vector of column names and want to select only those columns that match any of the names in the vector. For instance:

```{r message=FALSE}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
select(flights, any_of(vars))
```

This code selects columns with names "year," "month," "day," "dep_delay," and "arr_delay" from the **`flights`** dataframe.

The **`contains()`** helper function selects columns that contain a specified string in their names. By default, it's case-insensitive. For example:

```{r message=FALSE}
select(flights, contains("TIME"))
```

This code selects columns whose names contain the string "TIME," such as "dep_time" and "arr_time."

# 5.5.2 Exercises: Items 1 and 2

```{r message=FALSE}
flights_modified <- flights %>%
  mutate(
    dep_time_mins = (dep_time %/% 100) * 60 + dep_time %% 100,
    sched_dep_time_mins = (sched_dep_time %/% 100) * 60 + sched_dep_time %% 100)
```

In this code block:

1.  We take the **`flights`** dataset and pass it through the **`%>%`** operator.

2.  We use the **`mutate()`** function to add new columns to the data. We're creating two new columns: **`dep_time_mins`** and **`sched_dep_time_mins`**.

3.  For each of these columns, we're performing calculations to convert time values (stored in HHMM format) into minutes since midnight. We use the **`%/%`** operation to get the hours and **`%`** to get the minutes.

After executing this code block, the **`flights_modified`** dataset will contain the original columns along with the new **`dep_time_mins`** and **`sched_dep_time_mins`** columns, representing the departure times (scheduled and actual) in minutes since midnight.

```{r}
library(knitr)
kable(flights_modified[1:10,c(12,4,5,20,21)],caption = "SCHEDULED DEPARTURE TIME", align = "c")
```

```{r message=FALSE}
comparison_result <- flights_modified %>%
  mutate(arr_dep_time_diff = arr_time - dep_time_mins) %>%
  filter(!is.na(air_time) & !is.na(arr_dep_time_diff)) %>%
  select(air_time, arr_dep_time_diff)
```

In this second code block:

1.  We take the **`flights_modified`** dataset (the result of the previous block) and pass it through the **`%>%`** operator.

2.  We use the **`mutate()`** function to create a new column called **`arr_dep_time_diff`**. We're calculating the difference between the arrival time (**`arr_time`**) and the departure time in minutes since midnight (**`dep_time_mins`**).

3.  Next, we use **`filter()`** to remove rows where there are missing values in the **`air_time`** or **`arr_dep_time_diff`** columns.

4.  Finally, we use **`select()`** to choose only the **`air_time`** and **`arr_dep_time_diff`** columns.

```{r}
library(knitr)
kable(comparison_result[1:10,c(1,2)],caption = "COMPARISION OF ARRIVES AND DEPARTURES", align = "c")
```

# 5.6.7 Exercises: item 1

1.  **Median Arrival Delay**: Calculate the median arrival delay for the group of flights. This provides a central value that represents the typical delay experienced upon arrival.

2.  **Proportion of Flights with Specific Delays**: Determine the percentage of flights that arrive either 15 minutes early or 15 minutes late, 30 minutes early or 30 minutes late, or 2 hours late. This helps understand the distribution of different delay scenarios within the group.

3.  **Average Departure Delay**: Calculate the average departure delay for the group of flights. This gives insight into the average delay that occurs before a flight takes off.

4.  **Punctuality Percentage**: Calculate the percentage of flights that are punctual (no arrival delay) and compare it to the percentage of flights that are extremely delayed (2 hours late). This provides an understanding of how often flights are on time versus significantly delayed.

5.  **Arrival Delay Distribution**: Create a histogram or density plot showing the distribution of arrival delays across all flights. This visual representation helps identify common delay ranges and outliers.

**Question: What's More Important - Arrival Delay or Departure Delay?**

This question addresses whether arrival delay or departure delay has a greater impact on the overall flight experience. It depends on various factors. Arrival delay affects passengers' schedules, connecting flights, and ground transportation plans. Departure delay can impact scheduling and cause inconvenience, but it might be easier to manage in some cases compared to arrival delays that can ripple through subsequent plans.

Both aspects are important, but the significance may vary depending on passengers' priorities and the nature of their travel plans.

# 5.7.1 Exercises: item 2

```{r echo=FALSE}
worst_punctuality <- flights %>%
  group_by(tailnum) %>%
  summarize(
    total_flights = n(),
    punctual_flights = sum(arr_delay <= 0, na.rm = TRUE),
    punctuality_percentage = (punctual_flights / total_flights) * 100
  ) %>%
  arrange(punctuality_percentage) %>%
  filter(!is.na(punctuality_percentage))

worst_punctuality
```

In this code:

1.  We start with the **`flights`** dataset and use the **`%>%`** operator to chain subsequent operations.

2.  We group the data by aircraft tail number (**`tailnum`**). This means that subsequent calculations will be performed separately for each aircraft.

3.  We use the **`summarize()`** function to calculate summary statistics for each group (aircraft). Within the **`summarize()`** function:

    -   **`total_flights`** is calculated using the **`n()`** function, which gives the total number of flights for each aircraft.

    -   **`punctual_flights`** is calculated using the **`sum()`** function. It counts the number of flights where the arrival delay (**`arr_delay`**) is less than or equal to 0 (indicating on-time or early arrivals). The **`na.rm = TRUE`** argument handles missing values in the **`arr_delay`** column.

    -   **`punctuality_percentage`** is calculated as the ratio of punctual flights to total flights, multiplied by 100 to get the percentage.

4.  After the **`summarize()`** operation, we use **`arrange()`** to sort the groups (aircraft) based on their punctuality percentages in ascending order. This means aircraft with the lowest punctuality percentages will appear first.

5.  We use **`filter()`** to remove rows where the **`punctuality_percentage`** is not available (**`NA`**).

6.  The resulting dataset is assigned to **`worst_punctuality`**, and it contains aircraft tail numbers, the total number of flights, the number of punctual flights, and the corresponding punctuality percentages.

7.  Finally, we display the **`worst_punctuality`** dataset to see the tail numbers of aircraft with the lowest punctuality percentages.

```{r}
library(knitr)
kable(worst_punctuality[1:10,c(1,2,3,4)],caption = "WORST PUNCTUALITY TOP", align = "c")
```
